---
title: "RELIEF: integrative analysis"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

# Introduction

Here I'm using the new mitch package to compare methylation and RNA expression data from
the RELIEF patients.

Mitch: https://doi.org/10.1186/s12864-020-06856-9

The contrasts I'm looking at are:

1. Base-line versus post-op for all patients.

2. Low versus high CRP groups in post-op samples.

3. Low versus high CRP groups in base-line samples.

# Packages 

```{r,packages}
suppressPackageStartupMessages({
    library("gplots")
    library("mitch")
    library("kableExtra")
})
```

# Import read counts

Need to import the data and merge the RNA and array data.
I notices that only 11370 genes had promoter methylation values, while 21907 had RNA expression values.
We might need to look at updating the array annotation to include more genes as the array annotation is from ~2014 and might be missing a lot of ncRNA genes.

```{r,importdata}

# t0_v_pod_rna.tsv t0vPod_meth.tsv t0_v_pod
x <- read.table("t0_v_pod_rna.tsv",header=TRUE, sep="\t")
x$gene <- sapply(strsplit(rownames(x)," "),"[[",2)
x <- x[,c(4,ncol(x))]
x <- aggregate(. ~ gene, x, sum)
rownames(x) <- x$gene
x$gene=NULL

y <- read.csv("t0vPod_meth.tsv")
y$genename <- sapply(strsplit(y[,1],"\t"),"[[",1)
y$value <- as.numeric(sapply(strsplit(y[,1],"\t"),"[[",2))
y$t = NULL
row.names(y) <- y$genename
y$t = y$genename = NULL

z <- merge(x,y,by=0)
rownames(z) <- z$Row.names
z$Row.names = NULL
colnames(z) <- c("RNA","meth")
t0_v_pod <- z

# pod_crp_rna.tsv t1_crp_meth.tsv pod_crp
x <- read.table("pod_crp_rna.tsv",header=TRUE, sep="\t")
x$gene <- sapply(strsplit(rownames(x)," "),"[[",2)
x <- x[,c(4,ncol(x))]
x <- aggregate(. ~ gene, x, sum)
rownames(x) <- x$gene
x$gene=NULL

y <- read.csv("t1_crp_meth.tsv")
y$genename <- sapply(strsplit(y[,1],"\t"),"[[",1)
y$value <- as.numeric(sapply(strsplit(y[,1],"\t"),"[[",2))
y$t = NULL
row.names(y) <- y$genename
y$t = y$genename = NULL

z <- merge(x,y,by=0)
rownames(z) <- z$Row.names
z$Row.names = NULL
colnames(z) <- c("RNA","meth")
pod_crp <- z

# t0_crp_rna.tsv t0_crp_meth.tsv t0_crp
x <- read.table("t0_crp_rna.tsv",header=TRUE, sep="\t")
x$gene <- sapply(strsplit(rownames(x)," "),"[[",2)
x <- x[,c(4,ncol(x))]
x <- aggregate(. ~ gene, x, sum)
rownames(x) <- x$gene
x$gene=NULL

y <- read.csv("t0_crp_meth.tsv")
y$genename <- sapply(strsplit(y[,1],"\t"),"[[",1)
y$value <- as.numeric(sapply(strsplit(y[,1],"\t"),"[[",2))
y$t = NULL
row.names(y) <- y$genename
y$t = y$genename = NULL

z <- merge(x,y,by=0)
rownames(z) <- z$Row.names
z$Row.names = NULL
colnames(z) <- c("RNA","meth")
t0_crp <- z


```

## Obtaining reactome gene sets

```{r,reactome}

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

```

## Integrative pathway analysis

For each of the three contrast indicated above I have set up mitch analyses.

```{r,mitch}

# t0_v_pod
head(t0_v_pod) %>% kbl() %>% kable_paper("hover", full_width = F)
dim(t0_v_pod)

capture.output(
    res <- mitch_calc(x=t0_v_pod, genesets=genesets,priority="significance", resrows = 30)
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,30) %>% kbl() %>% kable_paper("hover", full_width = F)

unlink("int_t0_v_pod_sig.html")
capture.output(
    mitch_report(res, "int_t0_v_pod_sig.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res,"int_01_t0_v_pod_sig.pdf")

# new heatmap
colfunc <- colorRampPalette(c("blue", "white", "red"))
top <- res$enrichment_result[1:50,c(4,5)]
rownames(top) <- res$enrichment_result[1:50,1]
pdf("int_01_t0_v_pod_sig_heatmap.pdf",width=8,height=8)
heatmap.2( as.matrix(top), col=colfunc(25),scale="none",
 trace="none",margins = c(10,28), cexRow=.6, Colv=FALSE, dendrogram="row", cexCol=.8,  main="T0 v POD")
dev.off()

capture.output(
    res <- mitch_calc(x=t0_v_pod, genesets=genesets,priority="effect" , resrows = 30)
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,30) %>% kbl() %>% kable_paper("hover", full_width = F)

unlink("int_t0_v_pod_eff.html")
capture.output(
    mitch_report(res, "int_t0_v_pod_eff.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res,"int_01_t0_v_pod_eff.pdf")

top <- res$enrichment_result[1:50,c(4,5)]
rownames(top) <- res$enrichment_result[1:50,1]
pdf("int_01_t0_v_pod_eff_heatmap.pdf",width=8,height=8)
heatmap.2( as.matrix(top), col=colfunc(25),scale="none",
 trace="none",margins = c(10,28), cexRow=.6, Colv=FALSE, dendrogram="row", cexCol=.8,  main="T0 v POD")
dev.off()


# pod_crp
head(pod_crp) %>% kbl() %>% kable_paper("hover", full_width = F)
dim(pod_crp)

capture.output(
    res <- mitch_calc(x=pod_crp, genesets=genesets,priority="significance", resrows = 30)
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,30) %>% kbl() %>% kable_paper("hover", full_width = F)

unlink("int_pod_crp_sig.html")
capture.output(
    mitch_report(res, "int_pod_crp_sig.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res,"int_02_pod_crp_sig.pdf")

top <- res$enrichment_result[1:50,c(4,5)]
rownames(top) <- res$enrichment_result[1:50,1]
pdf("int_02_pod_crp_sig_heatmap.pdf",width=8,height=8)
heatmap.2( as.matrix(top), col=colfunc(25),scale="none",
 trace="none",margins = c(10,28), cexRow=.6, Colv=FALSE, dendrogram="row", cexCol=.8,  main="POD CRP")
dev.off()

capture.output(
    res <- mitch_calc(x=pod_crp, genesets=genesets,priority="effect", resrows = 30)
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,30) %>% kbl() %>% kable_paper("hover", full_width = F)

unlink("int_pod_crp_eff.html")
capture.output(
    mitch_report(res, "int_pod_crp_eff.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res,"int_02_pod_crp_eff.pdf")

top <- res$enrichment_result[1:50,c(4,5)]
rownames(top) <- res$enrichment_result[1:50,1]
pdf("int_02_pod_crp_eff_heatmap.pdf",width=8,height=8)
heatmap.2( as.matrix(top), col=colfunc(25),scale="none",
 trace="none",margins = c(10,28), cexRow=.6, Colv=FALSE, dendrogram="row", cexCol=.8,  main="POD CRP")
dev.off()


# t0_crp
head(t0_crp) %>% kbl() %>% kable_paper("hover", full_width = F)
dim(t0_crp)

capture.output(
    res <- mitch_calc(x=t0_crp, genesets=genesets,priority="significance", resrows = 30)
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,30) %>% kbl() %>% kable_paper("hover", full_width = F)

unlink("int_t0_crp_sig.html")
capture.output(
    mitch_report(res, "int_t0_crp_sig.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res,"int_03_t0_crp_sig.pdf")

top <- res$enrichment_result[1:50,c(4,5)]
rownames(top) <- res$enrichment_result[1:50,1]
pdf("int_03_t0_crp_sig_heatmap.pdf",width=8,height=8)
heatmap.2( as.matrix(top), col=colfunc(25),scale="none",
 trace="none",margins = c(10,28), cexRow=.6, Colv=FALSE, dendrogram="row", cexCol=.8,  main="t0 CRP")
dev.off()

capture.output(
    res <- mitch_calc(x=t0_crp, genesets=genesets,priority="effect", resrows = 30)
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,30) %>% kbl() %>% kable_paper("hover", full_width = F)

unlink("int_t0_crp_eff.html")
capture.output(
    mitch_report(res, "int_t0_crp_eff.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res,"int_03_t0_crp_eff.pdf")

top <- res$enrichment_result[1:50,c(4,5)]
rownames(top) <- res$enrichment_result[1:50,1]
pdf("int_03_t0_crp_eff_heatmap.pdf",width=8,height=8)
heatmap.2( as.matrix(top), col=colfunc(25),scale="none",
 trace="none",margins = c(10,28), cexRow=.6, Colv=FALSE, dendrogram="row", cexCol=.8,  main="t0 CRP")
dev.off()


```

## Four way mitch analysis


```{r, four_way}

colnames(t0_crp) <-  c("t0_crp_RNA", "t0_crp_meth" )

colnames( pod_crp ) <- c("pod_crp_RNA", "pod_crp_meth" )

xx <- merge(t0_crp,pod_crp,by=0)

rownames(xx) <- xx[,1]

xx[,1]=NULL

dim(xx)

head(xx) %>% kbl() %>% kable_paper("hover", full_width = F)

capture.output(
    res <- mitch_calc(x=xx, genesets=genesets,priority="significance", resrows = 30)
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

head(res$enrichment_result,30) %>% kbl() %>% kable_paper("hover", full_width = F)

unlink("int_4way_sig.html")
capture.output(
    mitch_report(res, "int_4way_sig.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res,"int_4way_sig.pdf")

top <- res$enrichment_result[1:50,c(4:7)]
rownames(top) <- res$enrichment_result[1:50,1]
pdf("int_4way_sig_heatmap.pdf",width=8,height=8)
heatmap.2( as.matrix(top), col=colfunc(25),scale="none",
 trace="none",margins = c(10,28), cexRow=.6, Colv=FALSE, dendrogram="row", cexCol=.8,  main="4way")
dev.off()




```




## Session information

```{r,sessioninfo}

sessionInfo()

```
