---
title: "RELIEF: RNA expression analysis"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

# Introduction


# Packages 

```{r,packages}

suppressPackageStartupMessages({
    library("zoo")
    library("tidyverse")
    library("reshape2")
    library("DESeq2")
    library("gplots")
    library("fgsea")
    library("MASS")
    library("mitch")
    library("eulerr")
    library("limma")
    library("topconfects")
})

```

# Import read counts

```{r,importdata}

tmp <- read.table("3col.tsv",header=F)
x <- as.matrix(acast(tmp, V2~V1, value.var="V3", fun.aggregate = sum))
x <- as.data.frame(x)
accession <- sapply((strsplit(rownames(x),"\\|")),"[[",2)
symbol<-sapply((strsplit(rownames(x),"\\|")),"[[",6)
x$geneid <- paste(accession,symbol)

xx <- aggregate(. ~ geneid,x,sum)
rownames(xx) <- xx$geneid
colnames <- gsub("T0R","T0",colnames(xx))
xx$geneid = NULL
xx <- round(xx)

```

# Curate the samplesheet

Here we're manipulating the sample sheet.
Making sure that the samplesheet matches the names of the data sets and vice versa.
Ensuring that each patient has a baseline and postop sample.

```{r,samplesheet}

samplesheet<-read.csv("samplesheet.tsv",sep="\t",stringsAsFactors=FALSE)
mysamples <- colnames(xx)
baseline <- mysamples[grep("T0",mysamples)]
postop <- mysamples[grep("POD",mysamples)]
baseline <- sapply(strsplit(baseline,"-"),"[[",1)
postop <- sapply(strsplit(postop,"-"),"[[",1)
mysamples <- intersect(baseline,postop)
samplesheet2 <- samplesheet
rownames(samplesheet2) <- paste(samplesheet2$RG_number,"T0",sep="-")
samplesheet3 <- samplesheet2
rownames(samplesheet2) <- paste(samplesheet2$RG_number,"T0",sep="-")
rownames(samplesheet3) <- paste(samplesheet2$RG_number,"POD",sep="-")
samplesheet2$timepoint <- 0       
samplesheet3$timepoint <- 1
samplesheet <- rbind(samplesheet2,samplesheet3)

# filter the data matrix
xx <- xx[,which(colnames(xx) %in% rownames(samplesheet))]
samplesheet <- samplesheet[which(rownames(samplesheet) %in% colnames(xx)),]
samplesheet <- samplesheet[order(rownames(samplesheet)),]

```

# Blood cell composition correction

Here the idea is to use the concept of Monaco et al, 2019 to deconvolute blood RNA-seq signature.

DOI:https://doi.org/10.1016/j.celrep.2019.01.041


```{r,blood}

xn <- xx
gt <- as.data.frame(sapply(strsplit(rownames(xn)," "),"[[",2) )
rownames(gt) <- rownames(xx)
colnames(gt) = "genesymbol"
gt$geneID <- rownames(xx)
blood <- read.table("https://raw.githubusercontent.com/giannimonaco/ABIS/master/data/sigmatrixRNAseq.txt")
blood2 <- merge(gt,blood,by.x="genesymbol",by.y=0)
blood2 <- blood2[which(!duplicated(blood2$genesymbol)),]
rownames(blood2) <- blood2$geneID
blood2 <- blood2[,c(3:ncol(blood2))]
genes <- intersect(rownames(xx), rownames(blood2))
dec <- apply(xx[genes, , drop=F], 2, function(x) coef(rlm( as.matrix(blood2[genes,]), x, maxit =100 ))) *100

dec <- t(dec/colSums(dec)*100)

dec <- signif(dec, 3)

cs <- colSums(dec)
order(-cs)

cv <- apply(dec,2,function(x) {sd(x)/mean(x)} )
cv[order(-cv)]

plot(cs,cv,main="cell abundance versus SD")
text(cs,cv+0.5,label=names(cs))

pdf("rnaseq_wcc_01_cscv.pdf")
plot(cs,cv,main="cell abundance versus SD")
text(cs,cv+0.5,label=names(cs))
dev.off()

colfunc <- colorRampPalette(c("blue", "white", "red"))

heatmap.2( dec, col=colfunc(25),scale="row",
 trace="none",margins = c(10,15), cexRow=.7, cexCol=.8,  main="cell type abundances")

pdf("rnaseq_wcc_02_cellheat.pdf")
heatmap.2( dec, col=colfunc(25),scale="row",
 trace="none",margins = c(10,15), cexRow=.7, cexCol=.8,  main="cell type abundances")
dev.off()

samplesheet <- cbind(samplesheet,dec)

dec_t0_lo <- as.data.frame(dec[which(rownames(dec) %in% rownames(samplesheet)[which(samplesheet$CrpGroup==0 & samplesheet$timepoint==0 )] ),])
dec_po_lo <- as.data.frame(dec[which(rownames(dec) %in% rownames(samplesheet)[which(samplesheet$CrpGroup==0 & samplesheet$timepoint==1 )] ),])
dec_t0_hi <- as.data.frame(dec[which(rownames(dec) %in% rownames(samplesheet)[which(samplesheet$CrpGroup==1 & samplesheet$timepoint==0 )] ),])
dec_po_hi <- as.data.frame(dec[which(rownames(dec) %in% rownames(samplesheet)[which(samplesheet$CrpGroup==1 & samplesheet$timepoint==1 )] ),])


boxplot(dec_t0_lo$Monocytes.C , dec_t0_hi$Monocytes.C, dec_po_lo$Monocytes.C,  dec_po_hi$Monocytes.C ,
  ylim=c(0,100),names=c("t0 lo","t0 hi","po lo","po hi"),main="Monocytes.C")
t.test(dec_t0_hi$Monocytes.C , dec_po_hi$Monocytes.C)

boxplot(dec_t0_lo$NK , dec_t0_hi$NK, dec_po_lo$NK,  dec_po_hi$NK ,
  ylim=c(0,15),names=c("t0 lo","t0 hi","po lo","po hi"),main="NK")
t.test(dec_t0_lo$NK , dec_po_lo$NK)

boxplot(dec_t0_lo$T.CD8.Memory , dec_t0_hi$T.CD8.Memory, dec_po_lo$T.CD8.Memory,  dec_po_hi$T.CD8.Memory ,
  ylim=c(0,40),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD8.Memory")

boxplot(dec_t0_lo$T.CD8.Memory , dec_t0_hi$T.CD8.Memory, dec_po_lo$T.CD8.Memory,  dec_po_hi$T.CD8.Memory ,
  ylim=c(0,40),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD8.Memory")

boxplot(dec_t0_lo$T.CD4.Naive , dec_t0_hi$T.CD4.Naive, dec_po_lo$T.CD4.Naive,  dec_po_hi$T.CD4.Naive ,
  ylim=c(0,25),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD4.Naive")

boxplot(dec_t0_lo$T.CD8.Naive , dec_t0_hi$T.CD8.Naive, dec_po_lo$T.CD8.Naive,  dec_po_hi$T.CD8.Naive ,
  ylim=c(0,15),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD8.Naive")

boxplot(dec_t0_lo$B.Naive , dec_t0_hi$B.Naive, dec_po_lo$B.Naive,  dec_po_hi$B.Naive ,
  ylim=c(0,10),names=c("t0 lo","t0 hi","po lo","po hi"),main="B.Naive")

boxplot(dec_t0_lo$T.CD4.Memory , dec_t0_hi$T.CD4.Memory, dec_po_lo$T.CD4.Memory,  dec_po_hi$T.CD4.Memory ,
  ylim=c(0,10),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD4.Memory")

boxplot(dec_t0_lo$MAIT , dec_t0_hi$MAIT, dec_po_lo$MAIT,  dec_po_hi$MAIT ,
  ylim=c(0,15),names=c("t0 lo","t0 hi","po lo","po hi"),main="MAIT")

boxplot(dec_t0_lo$T.gd.Vd2 , dec_t0_hi$T.gd.Vd2 , dec_po_lo$T.gd.Vd2 ,  dec_po_hi$T.gd.Vd2 ,
  ylim=c(0,5),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.gd.Vd2")

boxplot(dec_t0_lo$Neutrophils.LD , dec_t0_hi$Neutrophils.LD , dec_po_lo$Neutrophils.LD ,  dec_po_hi$Neutrophils.LD ,
  ylim=c(0,50),names=c("t0 lo","t0 hi","po lo","po hi"),main="Neutrophils.LD")

boxplot(dec_t0_lo$T.gd.non.Vd2 , dec_t0_hi$T.gd.non.Vd2 , dec_po_lo$T.gd.non.Vd2 ,  dec_po_hi$T.gd.non.Vd2 ,
  ylim=c(0,1),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.gd.non.Vd2")

boxplot(dec_t0_lo$Basophils.LD , dec_t0_hi$Basophils.LD , dec_po_lo$Basophils.LD ,  dec_po_hi$Basophils.LD ,
  ylim=c(0,4),names=c("t0 lo","t0 hi","po lo","po hi"),main="Basophils.LD")

boxplot(dec_t0_lo$Monocytes.NC.I , dec_t0_hi$Monocytes.NC.I , dec_po_lo$Monocytes.NC.I ,  dec_po_hi$Monocytes.NC.I ,
  ylim=c(0,20),names=c("t0 lo","t0 hi","po lo","po hi"),main="Monocytes.NC.I")

boxplot(dec_t0_lo$B.Memory , dec_t0_hi$B.Memory , dec_po_lo$B.Memory ,  dec_po_hi$B.Memory ,
  ylim=c(0,20),names=c("t0 lo","t0 hi","po lo","po hi"),main="B.Memory")

boxplot(dec_t0_lo$mDCs , dec_t0_hi$mDCs , dec_po_lo$mDCs ,  dec_po_hi$mDCs ,
  ylim=c(0,1),names=c("t0 lo","t0 hi","po lo","po hi"),main="mDCs")

boxplot(dec_t0_lo$pDCs , dec_t0_hi$pDCs , dec_po_lo$pDCs ,  dec_po_hi$pDCs ,
  ylim=c(0,1),names=c("t0 lo","t0 hi","po lo","po hi"),main="pDCs")

boxplot(dec_t0_lo$Plasmablasts , dec_t0_hi$Plasmablasts , dec_po_lo$Plasmablasts ,  dec_po_hi$Plasmablasts ,
  ylim=c(0,1),names=c("t0 lo","t0 hi","po lo","po hi"),main="Plasmablasts")

pdf("rnaseq_wcc_03_cellboxplots.pdf")

boxplot(dec_t0_lo$NK , dec_t0_hi$NK, dec_po_lo$NK,  dec_po_hi$NK ,
  ylim=c(0,15),names=c("t0 lo","t0 hi","po lo","po hi"),main="NK")
t.test(dec_t0_lo$NK , dec_po_lo$NK)

boxplot(dec_t0_lo$T.CD8.Memory , dec_t0_hi$T.CD8.Memory, dec_po_lo$T.CD8.Memory,  dec_po_hi$T.CD8.Memory ,
  ylim=c(0,40),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD8.Memory")

boxplot(dec_t0_lo$T.CD8.Memory , dec_t0_hi$T.CD8.Memory, dec_po_lo$T.CD8.Memory,  dec_po_hi$T.CD8.Memory ,
  ylim=c(0,40),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD8.Memory")

boxplot(dec_t0_lo$T.CD4.Naive , dec_t0_hi$T.CD4.Naive, dec_po_lo$T.CD4.Naive,  dec_po_hi$T.CD4.Naive ,
  ylim=c(0,25),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD4.Naive")

boxplot(dec_t0_lo$T.CD8.Naive , dec_t0_hi$T.CD8.Naive, dec_po_lo$T.CD8.Naive,  dec_po_hi$T.CD8.Naive ,
  ylim=c(0,15),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD8.Naive")

boxplot(dec_t0_lo$B.Naive , dec_t0_hi$B.Naive, dec_po_lo$B.Naive,  dec_po_hi$B.Naive ,
  ylim=c(0,10),names=c("t0 lo","t0 hi","po lo","po hi"),main="B.Naive")

boxplot(dec_t0_lo$T.CD4.Memory , dec_t0_hi$T.CD4.Memory, dec_po_lo$T.CD4.Memory,  dec_po_hi$T.CD4.Memory ,
  ylim=c(0,10),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.CD4.Memory")

boxplot(dec_t0_lo$MAIT , dec_t0_hi$MAIT, dec_po_lo$MAIT,  dec_po_hi$MAIT ,
  ylim=c(0,15),names=c("t0 lo","t0 hi","po lo","po hi"),main="MAIT")

boxplot(dec_t0_lo$T.gd.Vd2 , dec_t0_hi$T.gd.Vd2 , dec_po_lo$T.gd.Vd2 ,  dec_po_hi$T.gd.Vd2 ,
  ylim=c(0,5),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.gd.Vd2")

boxplot(dec_t0_lo$Neutrophils.LD , dec_t0_hi$Neutrophils.LD , dec_po_lo$Neutrophils.LD ,  dec_po_hi$Neutrophils.LD ,
  ylim=c(0,50),names=c("t0 lo","t0 hi","po lo","po hi"),main="Neutrophils.LD")

boxplot(dec_t0_lo$T.gd.non.Vd2 , dec_t0_hi$T.gd.non.Vd2 , dec_po_lo$T.gd.non.Vd2 ,  dec_po_hi$T.gd.non.Vd2 ,
  ylim=c(0,1),names=c("t0 lo","t0 hi","po lo","po hi"),main="T.gd.non.Vd2")

boxplot(dec_t0_lo$Basophils.LD , dec_t0_hi$Basophils.LD , dec_po_lo$Basophils.LD ,  dec_po_hi$Basophils.LD ,
  ylim=c(0,4),names=c("t0 lo","t0 hi","po lo","po hi"),main="Basophils.LD")

boxplot(dec_t0_lo$Monocytes.NC.I , dec_t0_hi$Monocytes.NC.I , dec_po_lo$Monocytes.NC.I ,  dec_po_hi$Monocytes.NC.I ,
  ylim=c(0,20),names=c("t0 lo","t0 hi","po lo","po hi"),main="Monocytes.NC.I")

boxplot(dec_t0_lo$B.Memory , dec_t0_hi$B.Memory , dec_po_lo$B.Memory ,  dec_po_hi$B.Memory ,
  ylim=c(0,20),names=c("t0 lo","t0 hi","po lo","po hi"),main="B.Memory")

boxplot(dec_t0_lo$mDCs , dec_t0_hi$mDCs , dec_po_lo$mDCs ,  dec_po_hi$mDCs ,
  ylim=c(0,1),names=c("t0 lo","t0 hi","po lo","po hi"),main="mDCs")

boxplot(dec_t0_lo$pDCs , dec_t0_hi$pDCs , dec_po_lo$pDCs ,  dec_po_hi$pDCs ,
  ylim=c(0,1),names=c("t0 lo","t0 hi","po lo","po hi"),main="pDCs")

boxplot(dec_t0_lo$Plasmablasts , dec_t0_hi$Plasmablasts , dec_po_lo$Plasmablasts ,  dec_po_hi$Plasmablasts ,
  ylim=c(0,1),names=c("t0 lo","t0 hi","po lo","po hi"),main="Plasmablasts")

dev.off()


```

# MDS analysis

```{r,mds}

plot(cmdscale(dist(t(xx))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p", col="gray", pch=19, cex.axis=1.3,cex.lab=1.3, bty='l')
text(cmdscale(dist(t(xx))), labels=colnames(xx),cex=1.3)
heatmap.2(cor(xx),trace="none",margins = c(10,10), main="pearson correlation heatmap")
heatmap.2(cor(xx,method="s"),trace="none",margins = c(10,10), main="spearman correlation heatmap")

pdf("rnaseq_wcc_04_mds.pdf")
plot(cmdscale(dist(t(xx))), xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p", col="gray", pch=19, cex.axis=1.3,cex.lab=1.3, bty='l')
text(cmdscale(dist(t(xx))), labels=colnames(xx),cex=1.3)
heatmap.2(cor(xx),trace="none",margins = c(10,10), main="pearson correlation heatmap")
heatmap.2(cor(xx,method="s"),trace="none",margins = c(10,10), main="spearman correlation heatmap")
dev.off()


```


# DE analysis baseline versus post-op

```{r,de1}

# filter genes with fewer than 10 reads per sample
NAME = "t0_v_pod"

# correct for cell types
y <- xx[which(rowSums(xx)/ncol(xx)>=(10)),]
dds <- DESeqDataSetFromMatrix(countData = y , colData = samplesheet, 
  design = ~ Sex + age + 
  Monocytes.C + NK + T.CD8.Memory + T.CD4.Naive + T.CD8.Naive + B.Naive + MAIT +
  Neutrophils.LD + Basophils.LD + Monocytes.NC.I + B.Memory + pDCs +
  timepoint )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <-cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
t0_v_pod <- dge
dge[1:10,1:6]

sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste(NAME, SIG , "DGEs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, 
  xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")	
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(samplesheet$timepoint)
colCols <- gsub("0","gray",colCols)
colCols <- gsub("1","black",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
  trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
#t0_v_pod <- dge

pdf("rnaseq_wcc_05_de01.pdf")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3,
  xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(samplesheet$timepoint)
colCols <- gsub("0","gray",colCols)
colCols <- gsub("1","black",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
  trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
dev.off()

```

# DE analysis post-op low versus high CRP

```{r,de2}

# filter genes with fewer than 10 reads per sample
NAME = "pod_crp"
ss <- subset(samplesheet,timepoint==1)
xxx <- xx[,colnames(xx) %in% rownames(ss)]
y <- xxx[which(rowSums(xxx)/ncol(xxx)>=(10)),]
dds <- DESeqDataSetFromMatrix(countData = y , colData = ss, design = ~ Sex + age + 
  Monocytes.C + NK + T.CD8.Memory + T.CD4.Naive + T.CD8.Naive + B.Naive + MAIT +
  Neutrophils.LD + Basophils.LD + Monocytes.NC.I + B.Memory + pDCs +
  CrpGroup )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
dge[1:10,1:6]
sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste(NAME, SIG , "DGEs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(ss$CrpGroup)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,  
  trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
pod_crp <- dge
write.table(dge,file="pod_crp_rna.tsv",sep="\t",quote=FALSE)

pdf("rnaseq_wcc_06_de02.pdf")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(ss$CrpGroup)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
  trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
pod_crp <- dge
write.table(dge,file="pod_crp_rna.tsv",sep="\t",quote=FALSE)
dev.off()

```

Run a ROC based on PCA distance classification.

```{r,de2_roc}

# library size normalised
yy <- log2( (y/colSums(y)*1000000) + 0.001 )
yy <- yy[,which(rowMeans(cor(yy))>0.75)]

# crp group
crpgrp <- ss$CrpGroup
names(crpgrp) <- rownames(ss)
crpgrp <- crpgrp[which(names(crpgrp) %in% colnames(yy))]


# confects
confects <- deseq2_confects(res)

# Limma p-values
ROCp <- function(ngenes){
  pnames <- rownames(dge)[1:ngenes]
  topgenes <- as.data.frame(yy[which(rownames(yy) %in% pnames),])
  topgenes$ctrl <- rowMeans(topgenes[,which(crpgrp==0)])
  topgenes$case <- rowMeans(topgenes[,which(crpgrp==1)])
  mds <- plotMDS(topgenes,ndim=2)
  mds <- mds$cmdscale.out
  hypotenuse <- function(x) {
    sqrt(sum(unlist(lapply(x, function(x) { x^2 })),na.rm = TRUE ))
  }
  ctrl_dist <- apply(mds[1:(nrow(mds)-2),],1,function(x) { 
    a2=(x[1] - mds[(nrow(mds)-1),1] )^2
    b2=(x[1] - mds[(nrow(mds)-1),2] )^2
    sqrt(a2+b2)
  })
  case_dist <- apply(mds[1:(nrow(mds)-2),],1,function(x) {
    a2=(x[1] - mds[nrow(mds),1] )^2
    b2=(x[1] - mds[nrow(mds),2] )^2
    sqrt(a2+b2)
  })
  df <- data.frame(ctrl_dist,case_dist)
  df$classification_case <- unname(unlist(apply(df,1,function(x) {
    if (x[2]<x[1]) {TRUE} else {FALSE}
  }))) *1
  df$case_conf <- df$ctrl_dist / ( df$ctrl_dist + df$case_dist )
  df$ctrl_conf <- df$case_dist / ( df$ctrl_dist + df$case_dist )
  df$conf <- unname(apply(df,1,function(x) {
    if (x[4]>x[5]) {x[4]} else {x[5]}
  }))
  df$crpgrp <- crpgrp
  df <- df[order(-df$conf),]
  df$correct <- unname(unlist(apply(df,1,function(x) { if (x[3] == x[7]) {1} else {0} } )))
  tpr <- cumsum(df$correct) / nrow(df)
  tpr <- c(tpr,1)
  fpr <- cumsum((df$correct -1)*-1) / nrow(df)
  fpr <- c(fpr,1)
  # AUC 
  id <- order(fpr)
  AUC=round(sum(diff(fpr[id])*rollmean(tpr[id],2)),3)
  HEADER=paste("AUC =",AUC)
  plot(fpr,tpr,type="l",xlim=c(0,1),ylim=c(0,1),xlab="FPR",ylab="TPR",main=paste(ngenes,"genes"))
  mtext(HEADER)
  AUC
}

# topconfect genes
ROCc <- function(ngenes){
  pnames <- confects$table$name[1:ngenes]
  #pnames <- rownames(dge)[1:ngenes]
  topgenes <- as.data.frame(yy[which(rownames(yy) %in% pnames),])
  topgenes$ctrl <- rowMeans(topgenes[,which(crpgrp==0)])
  topgenes$case <- rowMeans(topgenes[,which(crpgrp==1)])
  mds <- plotMDS(topgenes,ndim=2)
  mds <- mds$cmdscale.out
  hypotenuse <- function(x) {
    sqrt(sum(unlist(lapply(x, function(x) { x^2 })),na.rm = TRUE ))
  }
  ctrl_dist <- apply(mds[1:(nrow(mds)-2),],1,function(x) { 
    a2=(x[1] - mds[(nrow(mds)-1),1] )^2
    b2=(x[1] - mds[(nrow(mds)-1),2] )^2
    sqrt(a2+b2)
  })
  case_dist <- apply(mds[1:(nrow(mds)-2),],1,function(x) { 
    a2=(x[1] - mds[nrow(mds),1] )^2
    b2=(x[1] - mds[nrow(mds),2] )^2
    sqrt(a2+b2)
  })
  df <- data.frame(ctrl_dist,case_dist)
  df$classification_case <- unname(unlist(apply(df,1,function(x) {
    if (x[2]<x[1]) {TRUE} else {FALSE}
  }))) *1
  df$case_conf <- df$ctrl_dist / ( df$ctrl_dist + df$case_dist )
  df$ctrl_conf <- df$case_dist / ( df$ctrl_dist + df$case_dist )
  df$conf <- unname(apply(df,1,function(x) {
    if (x[4]>x[5]) {x[4]} else {x[5]}
  }))
  df$crpgrp <- crpgrp
  df <- df[order(-df$conf),]
  df$correct <- unname(unlist(apply(df,1,function(x) { if (x[3] == x[7]) {1} else {0} } )))
  tpr <- cumsum(df$correct) / nrow(df)
  tpr <- c(tpr,1)
  fpr <- cumsum((df$correct -1)*-1) / nrow(df)
  fpr <- c(fpr,1)
  # AUC 
  id <- order(fpr)
  AUC=round(sum(diff(fpr[id])*rollmean(tpr[id],2)),3)
  HEADER=paste("AUC =",AUC)
  plot(fpr,tpr,type="l",xlim=c(0,1),ylim=c(0,1),xlab="FPR",ylab="TPR",main=paste(ngenes,"genes"))
  mtext(HEADER)
  AUC
}

ng <- c(2,3,5,7,10,13,15,17,20,30,40,60,80,100,150,200,300,500)

p_roc <- lapply(ng,ROCp)
p_roc <- unlist(p_roc)

c_roc <- lapply(ng,ROCc)
c_roc <- unlist(c_roc)

```

Now need to summarise what is the optimal number of genes in the panel.

```{r,auc}

plot(ng,p_roc,type="b",ylim=c(0.7,1),col="blue",xlab="no. genes", ylab="AUC")
points(ng,c_roc,type="b",col="red")
mtext("blue: top p-value genes, red:top confect genes")
grid()

# zoom in
plot(ng,p_roc,type="b",xlim=c(0,100),col="blue",xlab="no. genes", ylab="AUC")
points(ng,c_roc,type="b",col="red")
mtext("blue: top p-value genes, red:top confect genes")
grid()


pdf("rnaseq_wcc_07_roc.pdf")

plot(ng,p_roc,type="b",ylim=c(0.7,1),col="blue",xlab="no. genes", ylab="AUC")
points(ng,c_roc,type="b",col="red")
mtext("blue: top p-value genes, red:top confect genes")
grid()

# zoom in
plot(ng,p_roc,type="b",xlim=c(0,100),col="blue",xlab="no. genes", ylab="AUC")
points(ng,c_roc,type="b",col="red")
mtext("blue: top p-value genes, red:top confect genes")
grid()

dev.off()


```

From this we can see that p-value genes show better discriminitory power compared to topconfect genes at least when the number of gene$
In the range of ~150 genes shows the best result for topconfect genes with AUC of ~0.95
I also ran the p_roc analysis with more dimensions but the result was the same:

2d:0.951 (15g)

3d:0.945 (15g)

4d:0.953 (15g)

5d:0.953 (15g)

Now let's have a look at the heatmap of the top 15 genes with the best discriminitory power.

```{r,hm}

colCols <- as.character(ss$CrpGroup)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:15,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
 trace="none",margins = c(10,15), cexRow=1, cexCol=.7, main="Top 15 genes")

pdf("rnaseq_wcc_08_roc-heat.pdf")

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:15,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
 trace="none",margins = c(10,15), cexRow=1, cexCol=.7, main="Top 15 genes")

dev.off()

```

# DE analysis baseline low versus high CRP

```{r,de3}

# filter genes with fewer than 10 reads per sample
NAME = "t0_crp"
ss <- subset(samplesheet,timepoint==0)
xxx <- xx[,colnames(xx) %in% rownames(ss)]
y <- xxx[which(rowSums(xxx)/ncol(xxx)>=(10)),]
dds <- DESeqDataSetFromMatrix(countData = y , colData = ss, design = ~ Sex + age +
  Monocytes.C + NK + T.CD8.Memory + T.CD4.Naive + T.CD8.Naive + B.Naive + MAIT +
  Neutrophils.LD + Basophils.LD + Monocytes.NC.I + B.Memory + pDCs +
  CrpGroup )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
dge[1:10,1:6]
sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste(NAME, SIG , "DGEs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(ss$CrpGroup)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,  
 trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
t0_crp <- dge
write.table(dge,file="t0_crp_rna.tsv",sep="\t",quote=FALSE)

pdf("rnaseq_wcc_09_de03.pdf")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(ss$CrpGroup)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
 trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
t0_crp <- dge
write.table(dge,file="t0_crp_rna.tsv",sep="\t",quote=FALSE)
dev.off()


```

# DE analysis baseline versus post-op (low CRP)

```{r,de4}

NAME = "t0_vs_pod_lowCRP"
ss <- subset(samplesheet,CrpGroup==0)
xxx <- xx[,colnames(xx) %in% rownames(ss)]
y <- xxx[which(rowSums(xxx)/ncol(xxx)>=(10)),]
dds <- DESeqDataSetFromMatrix(countData = y , colData = ss, design = ~ Sex + age +
  Monocytes.C + NK + T.CD8.Memory + T.CD4.Naive + T.CD8.Naive + B.Naive + MAIT +
  Neutrophils.LD + Basophils.LD + Monocytes.NC.I + B.Memory + pDCs +
  timepoint )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
dge[1:10,1:6]
sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste(NAME, SIG , "DGEs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(ss$timepoint)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
 trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
t0_vs_pod_lowCRP <- dge
write.table(dge,file="t0_vs_pod_lowCRP_rna.tsv",sep="\t",quote=FALSE)

t0_vs_pod_lowCRP_up <- rownames( subset(dge,padj<0.05 & log2FoldChange > 0 ) )
t0_vs_pod_lowCRP_dn <- rownames( subset(dge,padj<0.05 & log2FoldChange < 0 ) )


pdf("rnaseq_wcc_10_de04.pdf")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(ss$timepoint)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
 trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
dev.off()

```

# DE analysis baseline versus post-op (high CRP)

```{r,de5}

NAME = "t0_vs_pod_highCRP"
ss <- subset(samplesheet,CrpGroup==1)
xxx <- xx[,colnames(xx) %in% rownames(ss)]
y <- xxx[which(rowSums(xxx)/ncol(xxx)>=(10)),]
dds <- DESeqDataSetFromMatrix(countData = y , colData = ss, design = ~ Sex + age +
  Monocytes.C + NK + T.CD8.Memory + T.CD4.Naive + T.CD8.Naive + B.Naive + MAIT +
  Neutrophils.LD + Basophils.LD + Monocytes.NC.I + B.Memory + pDCs +
  timepoint )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
dge[1:10,1:6]
sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste(NAME, SIG , "DGEs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(ss$timepoint)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
 trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
t0_vs_pod_highCRP <- dge
write.table(dge,file="t0_vs_pod_highCRP_rna.tsv",sep="\t",quote=FALSE)

t0_vs_pod_highCRP_up <- rownames( subset(dge,padj<0.05 & log2FoldChange > 0 ) )
t0_vs_pod_highCRP_dn <- rownames( subset(dge,padj<0.05 & log2FoldChange < 0 ) )

pdf("rnaseq_wcc_11_de05.pdf")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3, xlab="log2 base mean",
  ylim=c(-3,3),ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 xlim=c(-3,3),xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- as.character(ss$timepoint)
colCols <- gsub("1","orange",colCols)
colCols <- gsub("0","yellow",colCols)

colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(  as.matrix(dge[1:50,c(7:ncol(dge))]), col=colfunc(25),scale="row",
  ColSideColors=colCols ,
 trace="none",margins = c(6,10), cexRow=.6, cexCol=.5, main="Top 50 genes")
dev.off()


```

# Compare low and high patterns

DE4 and DE5 will be compared.

```{r,venn1}

v1 <- list("low CRP up"=t0_vs_pod_lowCRP_up, 
    "low CRP dn"=t0_vs_pod_lowCRP_dn,
    "high CRP up"=t0_vs_pod_highCRP_up,
    "high CRP dn"=t0_vs_pod_highCRP_dn)

plot(euler(v1),quantities = TRUE, main="Gene overlap")

pdf("rnaseq_wcc_12_venn1.pdf")
plot(euler(v1),quantities = TRUE, main="Gene overlap")
dev.off()


```

Set up mitch data.

```{r,mitch_setup}

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

# set up the gene table
gt <- as.data.frame(sapply(strsplit(rownames(xx)," "),"[[",2))
gt$id <- rownames(xx)
colnames(gt) <- c("gene name","id")

```

Now mitch analysis with DE4 and DE5

```{r,mitch1}

l <- list("low CRP"=t0_vs_pod_lowCRP, "high CRP"=t0_vs_pod_highCRP)
y <- mitch_import( l , DEtype="DESeq2", geneTable=gt)

# prioritisation by statistical significance
res <- mitch_calc(y, genesets, priority="significance",resrows=50)
head(res$enrichment_result,20)

lo_up <- res$enrichment_result[which(p.adjust(res$enrichment_result$p.low.CRP)<0.05 & res$enrichment_result$s.low.CRP>0),1]
lo_dn <- res$enrichment_result[which(p.adjust(res$enrichment_result$p.low.CRP)<0.05 & res$enrichment_result$s.low.CRP<0),1]
hi_up <- res$enrichment_result[which(p.adjust(res$enrichment_result$p.high.CRP)<0.05 & res$enrichment_result$s.high.CRP>0),1]
hi_dn <- res$enrichment_result[which(p.adjust(res$enrichment_result$p.high.CRP)<0.05 & res$enrichment_result$s.high.CRP<0),1]

v1 <- list("low CRP up"=lo_up,
    "low CRP dn"=lo_dn,
    "high CRP up"=hi_up,
    "high CRP dn"=hi_dn)

plot(euler(v1),quantities = TRUE, main="Gene set overlap")

pdf("rnaseq_wcc_13_venn2.pdf")
plot(euler(v1),quantities = TRUE, main="Gene set overlap")
dev.off()

unlink("t0_v_pod_lohi_mitch_sig.html")
capture.output(
    mitch_report(res, "t0_v_pod_lohi_mitch_sig.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res, "t0_v_pod_lohi_mitch_sig.pdf")

# prioritisation by statistical significance
res <- mitch_calc(y, genesets, priority="effect",resrows=50)
head(res$enrichment_result,20)

unlink("t0_v_pod_lohi_mitch_eff.html")
capture.output(
    mitch_report(res, "t0_v_pod_lohi_mitch_eff.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res, "t0_v_pod_lohi_mitch_eff.pdf")

# prioritisation by discordance
res <- mitch_calc(y, genesets, priority="SD",resrows=50)
head(res$enrichment_result,20)

unlink("t0_v_pod_lohi_mitch_sd.html")
capture.output(
    mitch_report(res, "t0_v_pod_lohi_mitch_sd.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

mitch_plots(res, "t0_v_pod_lohi_mitch_sd.pdf")

```


## One dimensional enrichment analysis

Not correcting for blood compostion changes.

```{r,1dpw, message = FALSE}
# t0_v_pod
# baseline versus post-op
y <- mitch_import(t0_v_pod, DEtype="DESeq2", geneTable=gt)
res <- mitch_calc(y, genesets, priority="significance",resrows=100) 
head(res$enrichment_result,20)
unlink("t0_v_pod_mitch.html")
capture.output(
    mitch_report(res, "t0_v_pod_mitch.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

# pod_crp
# postop low versus high CRP
y <- mitch_import(pod_crp, DEtype="DESeq2", geneTable=gt)
res <- mitch_calc(y, genesets, priority="significance",resrows=100)
head(res$enrichment_result,20)
unlink("pod_crp_mitch.html")
capture.output(
    mitch_report(res, "pod_crp_mitch.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

# t0_crp
# baseline low versus high CRP
y <- mitch_import(t0_crp, DEtype="DESeq2", geneTable=gt)
res <- mitch_calc(y, genesets, priority="significance",resrows=100)
head(res$enrichment_result,20)
unlink("t0_crp_mitch.html")
capture.output(
    mitch_report(res, "t0_crp_mitch.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)


## prioritise by effect size

# t0_v_pod
# baseline versus post-op
y <- mitch_import(t0_v_pod, DEtype="DESeq2", geneTable=gt)
res <- mitch_calc(y, genesets, priority="effect",resrows=100)
head(res$enrichment_result,20)
unlink("t0_v_pod_mitche.html")
capture.output(
    mitch_report(res, "t0_v_pod_mitche.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)


# pod_crp
# postop low versus high CRP
y <- mitch_import(pod_crp, DEtype="DESeq2", geneTable=gt)
res <- mitch_calc(y, genesets, priority="effect",resrows=100)
head(res$enrichment_result,20)
unlink("pod_crp_mitche.html")
capture.output(
    mitch_report(res, "pod_crp_mitche.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)


# t0_crp
# baseline low versus high CRP
y <- mitch_import(t0_crp, DEtype="DESeq2", geneTable=gt)
res <- mitch_calc(y, genesets, priority="effect",resrows=100)
head(res$enrichment_result,20)
unlink("t0_crp_mitche.html")
capture.output(
    mitch_report(res, "t0_crp_mitche.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)


# 3D analysys significance
z <- list("t0_v_pod"=t0_v_pod, "pod_crp"=pod_crp, "t0_crp"=t0_crp)
zz <- mitch_import(z, DEtype="DESeq2", geneTable=gt)
res <- mitch_calc(zz, genesets, priority="significance",resrows=100)
head(res$enrichment_result,20)
unlink("3d_mitch.html")
capture.output(
    mitch_report(res, "3d_mitch.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

# 3D analysys effect size
res <- mitch_calc(zz, genesets, priority="effect",resrows=100)
head(res$enrichment_result,20)
unlink("3d_mitche.html")
capture.output(
    mitch_report(res, "3d_mitche.html")
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)

```

## Session information

```{r,session}

sessionInfo()
save.image("rnaseq_analysis_full.Rdata")

```

